---
title: "Untitled"
author: "Anthony Di Fiore"
date: "3/23/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro

Mark Grabowski

Felsenstein - Phylogenies and the Comparative Method

We are taking phylogeny into account, not controlling for it

Persistent challenges
- data are observational and many factor affect the data
See T Hansen 2013


data are correlated due to phylogeny and we must incoporate it

how to take phylo into account- it is often unclear what evolutionary models to use and parameters to estimate

comparative data can have as many or few independent data points, can't know which is the case without the phylogeny

OLS assumes that residuals of the model are normal, have same variance, and ARE INDEPENDENT of one another (residuals are correlated -- related species grouped together with respect to residuals)

Huey et al (2019) basically by assuming a model of evolution, one can account for evolutionary covariances between residuals and correct for non-independence

1 solution: given a phylo with known branch lengths known and how the traits evolved on each branch - Felsenstein showed how it was possibke to account for phylo relartionships - independent contasts

phylo "signal" asks if there is a relationship between the data and the phylogeny

old way of doing PCM - independence of the DATA is not the assumption of standard, no phylo regreassion... rather the assumption is that the residuals are independent

if the phylo signal in the response variable comes from the predictor, there's no need for correcting for phylo

ols assumes residuals are normally distributed, and are independent 

gls assumes residuals are normally distributed, but don't need to be independent

Brownian motion simulation

amount of change is proportional to the amount of time that has taken place times the scaling factor


Species covary in proportion to the amount of time they share on the phylogeny
- simple BM assumes one scale factor (sigma2)
- [Q: how do you estimate sigma?] can estimate it from the data

what is the variance of A and what is the variance of B?

        |-----------t2-------A
---t1---|
        |-----------t3-------B

var(A) = sigma2(t1+t2)
var(B) = sigma2(t1+t3)
        
how much time do A and B share on the phylogeny - t1

what is their covariance?

cov(A, B) = sigma2(t1)

can fill in var/cov matric

sigma2  times  [t1 + t2   t1   ]
               [t1        t1+t3]

we can use time and phylo to generate a var/cov matrix that says how the residuals are related to one another

what's the problem? it is that other evolutionary models besides Brownian motion are possible (this model does allow for difference scale parameters on different branches of the tree)

T Hansen - relevant major contributions are
- introduction of O-U models to macroevolutionary analysis
-also role of evolvabailty as a constraint
-importance of scale in biology (see Houle et al. 2011)

Hansen noted that PCMs concentrated on resolving a statisical question, but lost site of the biological problems being addressed

BM assumes species who have ancestors who deviated from the optimal relationship between the trait and environment...

i.e. BM is okay under certain conditions...  but not when traits are evolving in a correlated way

Under BM, there's no tendency for species to adapt to optimal states influences by other factors

Think about ADAPTIVE LANDSCAPES - see Simpson 1944, Lande 1979, 2007

As traits evolve, there's a climbing process

OU model and macroevolution

- introduced OU models specifically to test hypotheses of trait adaptation to different niches on a phylogeny

- the is a BM component to OU models

dy = - alpha(y- theta)dt + sigma * dw

first part is the deterministic pull of the trait towards a primary optimum and second part is non-directed stochastic change (akin to BM)

dy = change in trait over time

Hansen main contributions
-
- phylogenetic effects d/n equal phylogenetic inertia 

former is any relationship between a trait and a phylogeny (phylo correlations)

latter is correlated residuals

rather phylo effects = phylo inertia + phylogenetic patterns (this latter is where we can find signals of adaptation)

Model important point today

OU model should be viewed as a continuum
  - moving from nearly instantaneous adaptation toward the optima
  - to one where residuals follow...
  
Two major ways of doing PCM:
Hypothesis testing
- test specific hypotheses based on our understanding of biology

Phylogenetic natural history
See Uyeda et al. 2018 - explore patterns seen in the data and how that is related to phylogeny

## Hypothesis Testing
SLOUCH

"Stochastic Linear OU for Comparative Hypotheses"

flexible... can do PLGS-like analyses (which usually involves only BM)

also takes into account measurement error (which can bias parameter estimates)

based on ideas of Hansen 1997; see Hansen 2014

SLOUCH makes it possible to test adaptive hypotheses while accounting for species relationships

mvSLOUCH (Bartoszek et al. 2012) allows for multiple traits to adapt to each other and to fixed and randomly evolving predictor variables

(multiple response variables)

slouch allows you to account for measurement errors (if it can be estimated from the data)


```{r}
rm(list=ls()) #Clear the workspace
library(slouch)
library(phytools)
library(treeplyr)
options(scipen=999)
f <- "/Users/ad26693/Downloads/Datasets/BrBo_Grabowski_2016.csv"
t <- "/Users/ad26693/Downloads/Datasets/10KPrimateTree.tre"

BrBo.data<-read.csv(f) #Brain and Body data from Isler et al. including measurement error estimated in Grabowski et al. 2016

ss <- "/Users/ad26693/Downloads/Datasets/DeCasien SS Data.csv"
SS.data<-read.csv(ss) #Social system data from DeCasien et al. 20 17

gs <- "/Users/ad26693/Downloads/Datasets/Avg_Group_Size_DeCasien.csv"
Avg_GS.data<-read.csv(gs) #Group Size data from DeCasien e t al. 2017


phylo.tree<-read.tree(t)
str(phylo.tree)
head(phylo.tree$tip.label)
plot(phylo.tree) #Labels too big!
plot(phylo.tree,cex=0.15)
plot(phylo.tree,cex=0.15)
nodelabels(cex=0.25)
head(phylo.tree$edge)
#Drop tip - Homo sapiens
phylo.tree
 drop.tip(phylo.tree, c("Homo_sapiens"))
 
head(BrBo.data)

#Brain/Body/Diet Data
BrBo.data$N.ECV.species

BrBo.data<-mutate(BrBo.data,me.ECV.species=BrBo.data$predict.sp.ECV/BrBo.data$N.ECV.species) #Also from dpylr BrBo.data<-mutate(BrBo.data,me.BoM.species=BrBo.data$predict.sp.BoM/BrBo.data$N.body. mass.species)

phylo.tree <- multi2di(phylo.tree, random=FALSE) #Fix problems with bifurcating branc hes
phylo.tree$edge.length[phylo.tree$edge.length==0] <- .Machine$double.eps

max(branching.times(phylo.tree))
l.tree<-max(branching.times(phylo.tree))
phylo.tree$edge.length<-phylo.tree$edge.length/l.tree ## rescale tree to height 1
max(branching.times(phylo.tree))

missing <- phylo.tree$tip.label[!phylo.tree$tip.label %in% BrBo.data$Genus_Species]
head(missing)

BrBo.trdata <- make.treedata(phylo.tree, BrBo.data,name_column="Genus_Species")

BrBo.SS.data<-merge(BrBo.data,SS.data,by="Genus_Species")
dim(BrBo.SS.data)
BrBo.SS.AGS.data<-merge(BrBo.SS.data,Avg_GS.data,by="Genus_Species")
dim(BrBo.SS.AGS.data)

BrBo.SS.AGS.data <- make.treedata(phylo.tree, BrBo.SS.AGS.data,name_column="Genus_Species")
BrBo.SS.AGS.data<-filter(BrBo.SS.AGS.data,!is.na(ECV.species.mean)&!is.na(Body.mass.species.mean)&!is.na(Social.System)) #filter is from the dplyr paackage
dim(BrBo.SS.AGS.data$dat)

BrBo.SS.AGS.data<-filter(BrBo.SS.AGS.data, N.ECV.male>=1 & N.body.mass.male>=1 & N.ECV.female>=1 & N.body.mass.female>=1 & N.ECV.species>1 & N.body.mass.species>1)
dim(BrBo.SS.AGS.data$dat)

max.vy<-var(log(BrBo.SS.AGS.data$dat$Body.mass.species.mean),na.rm=TRUE)

max.hl<-max(branching.times(phylo.tree))
library(ape)
reconstruction.ss <- ace(as.factor(BrBo.SS.AGS.data$dat$Social.System), BrBo.SS.AGS.data$phy, type = "d")
head(reconstruction.ss$lik.anc)

internal.regimes.ss <- apply(reconstruction.ss$lik.anc, 1, function(e) colnames(reconstruction.ss$lik.anc)[which.max(e)])

library(ggsci)
#Set colors
mypal <- pal_aaas("default", alpha = 0.8)(4)
library("scales")
show_col(mypal)
SS.shifts.total<-unlist(list(as.factor(BrBo.SS.AGS.data$dat$Social.System),factor(internal.regimes.ss))) #First unlist the regimes and combine the regimes at the tips wit h the internal regimes
edge.regimes <- factor(SS.shifts.total[BrBo.SS.AGS.data$phy$edge[,2]]) #Get ggplot colors used for plot to make on tree
reg.colors<-mypal[unique(SS.shifts.total)]
print(reg.colors) #Order: Left-up,left-down,right-up,right-down levels(edge.regimes)
plot(BrBo.SS.AGS.data$phy,edge.color = reg.colors[edge.regimes], edge.width = 1, cex = 0.2)
quartz()

ECV.ps<-slouch.fit(phy = BrBo.SS.AGS.data$phy,
                   species = BrBo.SS.AGS.data$phy$tip.label,
                   response = log(BrBo.SS.AGS.data$dat$ECV.species.mean),
                   #mv.response=BrBo.SS.AGS.data$dat$me.ECV.species,
                   hl_values = seq(0.00001, max.hl*3, length.out = 50),
                   vy_values = seq(0.00001, max.vy*3, length.out = 50),
                   hillclimb = FALSE,convergence = 150,
                   lower = c(0.00001, 0.00001))
summary(ECV.ps)

```
Analysis 3

```{r}
BrBo.SS.AGS.data$phy$node.label <- internal.regimes.ss
ECV.SS.BoM.OU<-slouch.fit(phy = BrBo.SS.AGS.data$phy,
                          species = BrBo.SS.AGS.data$phy$tip.label,
                          response = log(BrBo.SS.AGS.data$dat$ECV.species.mean),
                          fixed.fact = as.factor(BrBo.SS.AGS.data$dat$Social.System),
                          direct.cov = log(BrBo.SS.AGS.data$dat$Body.mass.species.mean),
                          mv.response= BrBo.SS.AGS.data$dat$me.ECV.species, # meas error
                          mv.direct.cov = BrBo.SS.AGS.data$dat$me.BoM.species, # meas error
                          hl_values = seq(0.00001, max.hl*3, length.out = 50),
                          vy_values = seq(0.00001, max.vy, length.out = 50),
                          hillclimb = FALSE,convergence = 150,
                          lower = c(0.00001, 0.00001))
summary(ECV.SS.BoM.OU)
```

mvMorph

Cody Prang presents

check out Bulter and King 2014: Phylogenetic Comparative Analysis: A Modeling Approach for Adaptive Evolution

AIC is a measure of relative goodness of fit of models

need to calc log likelihood and number of model parameters

likelihood = prob of data given a model

mvMorph = souped up version of OUCH 
- enables one to fit multivariate evolutionary models to continuous data given a tree

See Clavel et al. 2015 in Methods in Ecol and Evolution

R packages: OUCH, mvSLOUCH, SURFACE, mvMORPH, L1OU, PhylogeneticEM

see Boettinger, Coop "Is your phylogeny informative? Measuring the power of comparative methods"

Example analysis is from Prang et al. 2021 Science Advances

```{r}
library(mvMORPH)
library(phytools)
library(picante)
library(MASS)
library(geomorph)
f <- "/Users/ad26693/Downloads/Datasets/data.csv"
tree_file <- "/Users/ad26693/Downloads/Datasets/tree.nex"
tree <- read.nexus(tree_file)
data <- read.csv(f)
data <- na.omit(data[,3:29])
names <- data$taxon
geomean <- as.matrix(apply(data[-1], 1, function(x) exp(mean(log(x)))))
data["geomean"] <- geomean
data.gm <- as.matrix(apply(data[-1], 2, function(x) log(x/data$geomean)))
data.gm <- data.gm[,-27]
data.gm <- as.data.frame(data.gm)
data.gm["taxon"] <- names
pca <- prcomp(data.gm[-27], scale = FALSE)
summary(pca)
plot(pca)
colors <- c(rep("dark grey", 67), # Humans
           rep("purple", 62), # Chimpanzees and bonobos
           rep("dark blue", 23), # Eastern gorillas
           rep("dark green", 36), # Western gorillas
           rep("orange", 33), # Orangutans
           rep("royal blue", 54), # Hylobatids
           rep("dark red", 83), # Old World monkeys
           rep("light green", 58), # New World monkeys
           rep("pink", 1), # Ardipithecus ramidus
           rep("gold", 5)) # Australopithecus & Homo fossils
pc.scores <- as.data.frame(pca$x)
eqscplot(matrix(c(pc.scores[,1], pc.scores[,2]), nrow=length(pc.scores[,1])), xlab = "PC1 (47%)", ylab = "PC2 (16%)", cex=1.5, col = colors, pch = 19)
eqscplot(matrix(c(pc.scores[,1], pc.scores[,3]), nrow=length(pc.scores[,1])), xlab = "PC1 (47%)", ylab = "PC3 (13%)", cex=1.5, col = colors, pch = 19)

pc.scores["taxon"] <- data$taxon
pc.scores.avg <- aggregate(pc.scores[, 1:3], list(pc.scores$taxon), mean)
row.names(pc.scores.avg) <- pc.scores.avg$Group.1
pc.scores.avg <- pc.scores.avg[,2:4]
comp.data <- match.phylo.data(tree, pc.scores.avg)
physignal(A = as.matrix(comp.data$data), phy = comp.data$phy, iter = 999) 
plot(physignal(A = as.matrix(comp.data$data), phy = comp.data$phy, iter = 999))
physignal(A = as.matrix(comp.data$data[1]), phy = comp.data$phy, iter=999) 
physignal(A = as.matrix(comp.data$data[2]), phy = comp.data$phy, iter=999)
physignal(A = as.matrix(comp.data$data[3]), phy = comp.data$phy, iter=999)

  H3.regimes <- as.vector(c(rep("arb",1),
                       rep("semiterr",1),
                       rep("arb",4),
                       rep("susp",5),
                       rep("arb",6),
                       rep("susp",1),
                       rep("terr",1),
                       rep("semiterr",1),
                       rep("terr",1),
                       rep("biped", 1),
                       rep("susp",5),
                       rep("semiterr",2),
                       rep("susp",3),
                       rep("arb",3),
                       rep("semiterr", 1),
                       rep("arb", 3),
                       rep("semiterr",3),
                       rep("terr",7),
                       rep("arb", 2),
                       rep("semiterr", 1),
                       rep("arb", 1),
                       rep("semiterr", 1),
                       rep("biped", 4)))

names(H3.regimes) <- comp.data$phy$tip.label
print(H3.regimes)

H4.regimes <- as.vector(c(rep("palm",6),
                       rep("susp",5),
                       rep("palm",6),
                       rep("susp",1),
                       rep("kw",3),
                       rep("biped", 1),
                       rep("susp",5),
                       rep("kw",2),
                       rep("susp",3),
                       rep("palm",3),
                       rep("dig", 1),
                       rep("palm", 3),
                       rep("dig",1),
                       rep("palm",1),
                       rep("dig", 8),
                       rep("palm", 4),
                       rep("kw",1),
                       rep("biped", 4)))
names(H4.regimes) <- comp.data$phy$tip.label
print(H4.regimes)

H3.tree <- make.simmap(comp.data$phy, H3.regimes, model="ER", nsim=1)
plot(H3.tree, fsize = 0.5)

H4.tree <- make.simmap(comp.data$phy, H4.regimes, model="ER", nsims=1)
plot(H4.tree, fsize = 0.5)

H1.BM <- mvBM(tree = comp.data$phy,
              data = comp.data$data,
              model = "BM1",
              param = list(root = FALSE))
# H2. Single optimum Ornstein-Uhlenbeck (OU)
H2.OU1 <- mvOU(tree = comp.data$phy, data = comp.data$data, model = "OU1", param = list(root = FALSE))
# H3. Multi-optimum OU model
H3.OUM <- mvOU(tree = H3.tree, data = comp.data$data, model = "OUM", param = list(root = FALSE))
# H4. Multi-optimum OU model
H4.OUM <- mvOU(tree = H4.tree, data = comp.data$data, model = "OUM", param = list(root = FALSE))

AICc <- list(H1.BM$AICc, H2.OU1$AICc, H3.OUM$AICc, H4.OUM$AICc)
names(AICc) <- c("H1.BM", "H2.OU1", "H3.OUM", "H4.OUM")
print(AICc)
print(H4.OUM)
halflife(H2.OU1)
halflife(H3.OUM)
halflife(H4.OUM)
halflife(H1.BM)

fitA <- mvBM(tree = comp.data$phy, data = comp.data$data, model = "BM1", param = list (root = FALSE))
fitB <- mvOU(tree = comp.data$phy, data = comp.data$data, model = "OU1", param = list (root = FALSE))
lr_bmvou1 <- -2*(logLik(fitA) - logLik(fitB)) # Likelihood ratio statistic print(lr_bmvou1)
nsims <- 10
A_sims <- simulate(fitA, tree = comp.data$phy, nsims) # data simulated under BM
B_sims <- simulate(fitB, tree = comp.data$phy, nsims) # data simulated under OU1
AA <- lapply(1:nsims, function(i){
  mvBM(tree = comp.data$phy, data = A_sims[[i]], model="BM1", param = list(root = FALSE))})
# fit OU1 to data simulated under BM
AB <- lapply(1:nsims, function(i){
  mvOU(tree = comp.data$phy, data = A_sims[[i]], model="OU1", param = list(root = FALSE))})

# fit BM to data simulated under OU
BA <- lapply(1:nsims, function(i){
  mvBM(tree = comp.data$phy, data =
B_sims[[i]], model="BM1", param = list(root = FALSE))})

# fit OU1 to data simulated under OU
BB <- lapply(1:nsims, function(i){
mvOU(tree = comp.data$phy, data = B_sims[[i]], model="OU1", param = list(root = FALSE))})
```

## Phylogenetic Natural Hostoy
bayou (Mark Grabowski)

```{r}
closeAllConnections()
detach("package:treeplyr", unload = TRUE)
rm(list=ls()) #Clear the workspace
par(mfrow=c(1,1))
library(usethis)
usethis::edit_r_environ()
#R_MAX_VSIZE=100Gb

library(bayou)
#library(phytools)
#library(treeplyr)
library(dplyr)
#library(doParallel)
library(doParallel)
no_cores <- detectCores()
registerDoParallel(cores=no_cores)
options(scipen=999)

f <- "/Users/ad26693/Downloads/Datasets/BrBo_Grabowski_2016.csv"
t <- "/Users/ad26693/Downloads/Datasets/10KPrimateTree.tre"
phylo.tree<-read.tree(t) #10K Trees tree
BrBo.data<-read.csv(f)

max(branching.times(phylo.tree))
l.tree<-max(branching.times(phylo.tree))
phylo.tree$edge.length<-phylo.tree$edge.length/l.tree
max(branching.times(phylo.tree))
phylo.tree <- multi2di(phylo.tree, random=FALSE)#Fix problems with bifurcating branch es
phylo.tree$edge.length[phylo.tree$edge.length==0] <- .Machine$double.eps
phylo.tree<-reorder(phylo.tree,order = "postorder")

BrBo.data$N.ECV.species
BrBo.data<-mutate(BrBo.data,me.ECV.species=BrBo.data$predict.sp.ECV/BrBo.data$N.ECV.species)
BrBo.data<-mutate(BrBo.data,me.BoM.species=BrBo.data$predict.sp.BoM/BrBo.data$N.body.mass.species)

library(treeplyr)
BrBo.trdata <- make.treedata(phylo.tree, BrBo.data,name_column="Genus_Species")
BrBo.trdata<-filter(BrBo.trdata,!is.na(ECV.species.mean)&!is.na(Body.mass.species.mean))
dim(BrBo.trdata$dat)
BrBo.trdata<-filter(BrBo.trdata, N.ECV.male>=1 & N.body.mass.male>=1 & N.ECV.female>= 1 & N.body.mass.female>=1 & N.ECV.species>1 & N.body.mass.species>1)

log.ECV.dat<-log(BrBo.trdata[['ECV.species.mean']])
me.ECV.species<-BrBo.trdata[['me.ECV.species']]

log.BoM.dat<-log(BrBo.trdata[['Body.mass.species.mean']])
log.BoM.dat<-as.matrix(log.BoM.dat)
colnames(log.BoM.dat)<-"log.BoM.dat"
log.BoM.dat.vis<-log(BrBo.trdata[['Body.mass.species.mean']])
me.BoM.species<-BrBo.trdata[['me.BoM.species']]
phylo.tree<-BrBo.trdata$phy

phenogram(phylo.tree, log.ECV.dat, spread.labels=FALSE, fsize=0.4)
phenogram(phylo.tree, log.BoM.dat.vis, spread.labels=FALSE, fsize=0.4)
```

check out Uyeda github site

SURFACE
